---
title: "Climate Change Module"
author: 'Riley Spoenlein'
format:
  html:
    embed-resources: true
---

```{r message=FALSE}
library(tidyverse)
```

## Warm-up: Examining CO2 trends in R

- Example from <http://climate.nasa.gov/vital-signs/carbon-dioxide/>
- Raw data from <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt>

In 1958, Dr. Charles David Keeling (1928-2005), a scientist at Scripps Institute of Oceanography, began collecting data on atmospheric CO2 concentration at the Mauna Loa Observatory located in Hawaii. This dataset allowed us to understand the degree to which climate change is human-caused through our burning of fossil fuels and release of CO2 into the atmosphere. Due to his scientific achievements, Dr. Keeling was awarded the National Medal of Science by President George W. Bush in 2002. This is the highest award for lifetime scientific achievement that can be granted in the U.S. Today, you get to analyze this same dataset, except that you have more data than was available to Dr. Keeling and his colleagues because your dataset extends up to the current time.

To read the code, you will use a new function called `read_table`.  It is similar to `read_csv` except it looks for spaces between column entries rather than commas (remember that csv stands for comma-separated values).  Others include `read_tsv`, which uses tabs to separate entries.  You can discover the separation type by putting <https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt> into your web browser and examining the file.  The code also includes an argument for comment, which denotes the character used to define a line in the file as being a comment rather than data or a header.  The file also reveals that the column headers are on lines with a comment character, so they won't be read.  You will use the argument `col_names` to define the headers manually.  Finally, `NA` values are defined by the numbers -1 and -99, so they need to be defined using the `na` argument (otherwise, they will be read as numeric values).

```{r message=FALSE}

co2 <-  read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt", 
                  comment="#",
                  col_names = c("year", "month", "decimal_date",
                                "monthly_average",
                                "deseasonalized", "days", "sd_days",
                                "unc_month_mean"),
                  na = c("-1", "-99.99"))
co2
```


```{r}
ggplot(co2, aes(x = decimal_date, y = monthly_average)) + 
  geom_line() + 
  geom_line(aes(y = deseasonalized), color = "blue") +
  labs(x = "Year", y = "CO2 concentration (ppm)")
```

**Question 1:**

Describe the overall trend in the CO2 data.

**Answer 1:** The trend we can see in the above plot is a steady increase in CO2 concentration. It seems to be increasing more drastically in the last 15(ish) years.

**Question 2:**

How does CO2 vary within a year?  What month is it at max?  Min?  What might explain this sub-annual pattern? (you will need to write code and make a plot to answer this question)

```{r}
# Group by month and calculate the average monthly CO2 concentration
monthly_pattern <- co2 |>
  group_by(month) |>
  summarize(average_co2 = mean(monthly_average, na.rm = TRUE))

monthly_pattern$month <- factor(month.abb[monthly_pattern$month], levels = month.abb)

# Plot the average CO2 concentration by month
ggplot(monthly_pattern, aes(x = month, y = average_co2, group = 1)) +
  geom_line() +
  geom_point() +
  labs(x = "Month", y = "Average CO2 Concentration (ppm)",
       title = "Average Monthly CO2 Levels (All Years)")

```


**Answer 2:** We can see that the max levels are around May and lowest in Sep/Oct. CO2 increases at the end of fall and then drastically decreases at the end of spring. This could indicate that once the plants have grown (and are going through photosynthesis), they take in CO2 and cause the levels to drop in the summer. 



## Global Temperature Data

Current climate change affects many aspects of the environment, with socio-economic consequences. For example, a warmer climate can allow new diseases to be introduced and persist (e.g., West Nile became established in the United States after an unusually warm winter, which allowed the mosquitoes that carry the virus to survive and spread). We are concerned not only with the actual temperature but also with the rate at which it changes. Rapid changes make it more likely that species cannot adapt and will go extinct. 

Each of the most recent years has been the warmest on record. In this section, we will analyze global mean temperature data.

Data from: <https://climate.nasa.gov/vital-signs/global-temperature/>

**Question 3:**

Describe the data set to the best of your ability, given the documentation provided. 

- Where is the data from?
- Describe the class of each column and what units it is measured in. 
- What is the difference between "global temperature" and "global temperature anomaly"? 

**Answer 3:** This data is from NASA, which we can generally assume is reliable. The year (integer) column goes 1880-present. No_Smoothing (numeric) is the global temperature anomaly for the year in Celsius. Lowess(5) is also numeric and the smoothed version of global temperature anomaly. The temperature anomaly is the difference between the temperature and baseline average.   


**Question 4:**

Construct the necessary R code to import and prepare for plotting the following data set: <https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt>

You'll need to determine the file's delimiter and any comments or skips. You will also need to be sure that you have column names. You will not directly use the code above to read the CO2 data, but that code provides helpful tips for reading the data.

**Answer 4:**

```{r}
# Read the temperature data with proper cleaning
temp_data <- read_table(
  "https://data.giss.nasa.gov/gistemp/graphs/graph_data/Global_Mean_Estimates_based_on_Land_and_Ocean_Data/graph.txt",
  comment = "#",
  col_names = c("year", "no_smoothing", "lowess_smoothing")
) |>
  # Clean up any non-numeric rows caused by separator lines
  mutate(across(c(year, no_smoothing, lowess_smoothing), as.numeric)) |>
  drop_na() # Remove any rows with NA from separator lines

temp_data
```

**Question 5:**

Plot the trend in global mean temperatures over time.  Describe what you see in the plot and how you interpret the patterns you observe.

**Answer 5:**
```{r}
# Plot the global temperature anomalies over time
ggplot(temp_data, aes(x = year)) +
  geom_line(aes(y = no_smoothing), color = "red", alpha = 0.5) +
  geom_line(aes(y = lowess_smoothing), color = "blue", size = 1) +
  labs(
    x = "Year",
    y = "Temperature Anomaly (°C)",
    title = "Global Mean Temperature Anomalies (1880 - Present)",
    caption = "Data Source: NASA GISS Land-Ocean Temperature Index"
  ) +
  theme_minimal()
```
We see that similarly to the CO2 levels, global temperatures have greatly increased the past century. 

## Evaluating the evidence for a "Pause" in warming?

The [2013 IPCC Report](https://www.ipcc.ch/pdf/assessment-report/ar5/wg1/WG1AR5_SummaryVolume_FINAL.pdf) included a tentative observation of a "much smaller increasing trend" in global mean temperatures since 1998 than was observed previously.  This led to much discussion in the media about the existence of a "Pause" or "Hiatus" in global warming rates, as well as much research looking into where the extra heat could have gone.  (Examples discussing this question include articles in [The Guardian](http://www.theguardian.com/environment/2015/jun/04/global-warming-hasnt-paused-study-finds), [BBC News](http://www.bbc.com/news/science-environment-28870988), and [Wikipedia](https://en.wikipedia.org/wiki/Global_warming_hiatus)). 

You will use rolling averages to help you explore the evidence for a pause. Since you have not been provided instructions for calculating rolling means, the learning objective of this question is to practice finding the solution.

**Question 6:**

Use a search engine (e.g., Google) or a chat LLM to find out how to calculate a rolling average in R. What search term or chat LLM prompt did you use?  What website or answer from the chat LLM did you end up using?  How much did you need to modify the code from the website or chat LLM to answer the question successfully?

**Answer 6:**
The base code for this operation is:

library(zoo)
rolling_avg <- rollmean(data$temperature, k = 5, fill = NA)

I used chatgpt to find this function of the zoo library. The prompt I used was "how can I calculate a rolling average of global temperature in R". I would need to modify the variable names, ect. 

**Question 7:**

- What is the meaning of "5-year average" vs. "annual average"?
- Create a data frame from the annual temperature anomaly data (from Question 4) with three new columns: 5-year running averages, 10-year running averages, and 20-year running averages.

**Answer 7:** An annual average is the temperature anomaly for one year. The 5-year average smooths year-to-year fluctuations by looking at the average temperature anomaly over 5-year period. 

```{r}
library(zoo)

temp_data <- temp_data |>
  mutate(
    avg_5yr = rollmean(no_smoothing, k = 5, fill = NA, align = "center"),
    avg_10yr = rollmean(no_smoothing, k = 10, fill = NA, align = "center"),
    avg_20yr = rollmean(no_smoothing, k = 20, fill = NA, align = "center")
  )

temp_data
```

**Question 8:**

Plot the different averages on the *same plot* and describe what differences you see and why.  

**Answer 8:**
```{r}
ggplot(temp_data, aes(x = year)) +
  geom_line(aes(y = no_smoothing), color = "gray", alpha = 0.5) +
  geom_line(aes(y = avg_5yr), color = "blue", size = 1) +
  geom_line(aes(y = avg_10yr), color = "green", size = 1) +
  geom_line(aes(y = avg_20yr), color = "red", size = 1) +
  labs(
    x = "Year",
    y = "Temperature Anomaly (°C)",
    title = "Annual and Rolling Averages of Global Temperature Anomalies",
    caption = "Gray = Annual, Blue = 5-year, Green = 10-year, Red = 20-year"
  ) +
  theme_minimal()

```

It is clear that the higher year running averages have a smoother trend as expected. The annual average varies a lot year to year, while the 20-year running average gives us a good idea of the overall change a bit clearer. 


**Question 9:**

By examining the data here, what evidence do you find or not find for such a pause? 

**Answer 9:** It is possible to observe a little decrease/leveling around this time in the 5-year running average (blue line). However, it is not at all present in the other running averages. It is likely that this was simply a short term fluctuation, rather than a "pause". 

## Longer term trends in CO2 Records

When analyzing Earth’s climate, it is important to remember that Earth is 4.54 billion years old. Our analyses so far have only looked at recent history. How can we compare the recent data to prehistoric times? Are the current rates of change similar or different from those the earth has experienced in the past? To explore this, we can use data from ice cores drilled at the poles. 

Hundreds of ice cores have been extracted from polar ice because they contain valuable data on atmospheric chemistry over pre-historic times. These valuable data exist in tiny air bubbles trapped in the ice. These air bubbles contain the same gases in the same ratios as the atmosphere at the time when the ice formed. The data you will analyze today are from ice cores extracted from the Vostok research station in Antarctica. As you have probably assumed, the depth of the ice core is related to how old the ice is; deep ice is older. There are two other variables that you will be analyzing from the ice cores.  You will analyze CO2 concentration, which has been measured from air bubbles trapped in the ice. We can use these data to see what rates of change were like during this pre-historic period, during which human activity was minimal. 

[Ice core data](https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009):

Vostok Core, back to 400,000 years before the present day 

- Description of data set: <https://data.ess-dive.lbl.gov/view/doi%3A10.3334%2FCDIAC%2FATG.009>
- data: <https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542>

You will use this code to download the data to your computer.

```{r}
download.file("https://data.ess-dive.lbl.gov/catalog/d1/mn/v2/object/ess-dive-457358fdc81d3a5-20180726T203952542",
              destfile = "vostok.icecore.co2")
```

You can click on the file in your Files pane to view it before reading into R.


**Question 10:**

The broad question is: how do recent atmospheric CO2 levels compare to historical levels?

Your answer to Question 10 is going to be a mix of code chunks and text that you put below in "Answer 10:"

- Describe the data set: what are the columns and units? Where do the numbers come from? 
- What is the temporal resolution of the data? 
- Read in and prepare data for analysis.   
- Reverse the ordering to create a chronological record so that each measurement is associcated with calender year.
- Plot data.  
- Combine this time series with the Mauna Loa data (the Warm-up exercise in this assignment).  You will use "bind_rows()".
- Plot the combined data. (the most recent time period must be on the right side of the plot).        
- Describe your conclusions to the question "How do recent atmospheric CO2 levels compare to historical levels?" using your plot as supporting evidence.   

**Answer 10:**

```{r}
# Description of the dataset:
# - depth_m: Depth in the ice core (meters).
# - ice_age_bp: Age of the ice in years before present (up t0 1950).
# - air_age_bp: Age of the air bubbles in years before present (up to 1950).
# - co2_ppm: Atmospheric CO2 concentration in parts per million (ppm), extracted from trapped air bubbles.

# Temporal Resolution:
# - The data is irregularly spaced, with intervals of hundreds to a few thousand years.
# - The resolution is higher in more recent layers and lower in older layers.

# Read the data (skip 21 lines, set column names manually)
vostok_data <- read_table(
  "vostok.icecore.co2",
  skip = 21,
  col_names = c("depth_m", "ice_age_bp", "air_age_bp", "co2_ppm")
)


# Prepare the data:
# Convert "years before present" to calendar year (1950 is "present")
vostok_data <- vostok_data |>
  mutate(calendar_year = 1950 - air_age_bp) |>
  arrange(calendar_year)

# Plot Vostok ice core CO2 data
ggplot(vostok_data, aes(x = calendar_year, y = co2_ppm)) +
  geom_line(color = "blue") +
  labs(
    title = "Vostok Ice Core CO2 Levels (Past 400,000 Years)",
    x = "Year",
    y = "CO2 Concentration (ppm)"
  ) +
  theme_minimal()

# Read Mauna Loa CO2 data from the warm-up exercise
co2 <- read_table("https://gml.noaa.gov/webdata/ccgg/trends/co2/co2_mm_mlo.txt",
                  comment = "#",
                  col_names = c("year", "month", "decimal_date",
                                "monthly_average", "deseasonalized",
                                "days", "sd_days", "unc_month_mean"),
                  na = c("-1", "-99.99"))

# Process Mauna Loa data (filter NA, group by year)
co2_recent <- co2 |>
  filter(!is.na(monthly_average)) |>
  group_by(year) |>
  summarize(co2_ppm = mean(monthly_average, na.rm = TRUE)) |>
  rename(calendar_year = year)

# Combine Vostok and Mauna Loa Data
combined_co2 <- bind_rows(
  vostok_data |>
    select(calendar_year, co2_ppm),
  co2_recent
)

# Plot combined CO2 data
ggplot(combined_co2, aes(x = calendar_year, y = co2_ppm)) +
  geom_line(color = "black") +
  labs(
    title = "CO2 Concentrations: Vostok Ice Core vs. Mauna Loa (Modern)",
    x = "Year",
    y = "CO2 Concentration (ppm)"
  ) +
  theme_minimal()
```
This graph shows pretty consistent fluctuations of CO2 concentration over time. These seem to be generally between 100 and 300 ppm. However, we see that recently it has spiked over 400 which is not consistent with past numbers.
# Render and committing

Remember to Render your document as HTML and comment+push to GitHub your code and rendered HTML that was created when you knitted the document.  Your GitHub repository should have multiple commits with informative commit messages.

# Attribution

Include citation of any AI-generated assistance or discussion with classmates (per policy in syllabus). Proper documentation of AI-generated assistance includes the prompt, the source (e.g., ChatGPT), and the significant parts of the response.  Proper documentation of discussion with classmates include listing their names and the components discussed.  

I used chatgpt to find rollmean as stated above.

